### 2014-12-04 text/Chapter02.tex
Proofreading (Chapter 2)

\noindent
Without providing much details, we note that well known examples of monads include the partiality
monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in 
Haskell, list monad ($\mtyp{}{\alpha} = \mu \gamma.1 + (\alpha \times \gamma)$) and other.
In programming language semantics, monads can be used in two distinct ways.

@@

In this thesis, we generally follow the first approach -- this means that we work with an existing
programming language without needing to add additional constructs corresponding to the primitives
of our semantics (the alternative is discussed in Section~\ref{sec:unified-meta}). To clarify the 
difference, the following two sections show a minimal example of both formal systems. We follow 
Moggi and start with language where judgements have the form $x\!:\!\alpha \vdash e : \beta$ with 
exactly one variable\footnote{This simplifies the examples as we do not need \emph{strong} monad, 
but that is an orthogonal issue to the distinction between language semantics and meta-language.}.

@@

\paragraph{Language semantics.}
To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
$\ctyp{}{\alpha} = \mu \gamma.\alpha + (\alpha \times \gamma)$. A value of the type is either
the last element $\alpha$ or an element followed by another non-empty list $\alpha \times \gamma$
(consisting of the head $\alpha$ and the tail $\gamma$). Note that the list must be non-empty, 
otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In 
the following, we write $(l_1, \ldots, l_n)$ for a list of $n$ elements:
%
\begin{equation*}
\begin{array}{rcl}
\ident{counit}~(l_1, \ldots, l_n) &=& l_1\\
\ident{cobind}~f~(l_1, \ldots, l_n) &=& (f (l_1, \ldots, l_n), f (l_2, \ldots, l_n), \ldots, f (l_n))
\end{array}
\end{equation*}
%
The \ident{counit} operation returns the current (first) element of the (non-empty) list.
The \ident{cobind} operation creates a new list by applying the context-dependent function $f$
to the entire list, to the suffix of the list, to the suffix of the suffix and so on. Interestingly,
it preserves the \emph{shape} of the list as it turns a list of $n$ elements into another list
of $n$ elements.

@@

\paragraph{Meta-language interpretation.} ~To briefly demonstrate the app\-roach that employs comonads
as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
Nanevski et al. We do not attempt to provide precise overview of their work. The main purpose 
of our discussion is to provide a different intuition behind comonads, and to give an example
of a language that includes comonad as a type constructor, together with language primitives
corresponding to comonadic operations\footnote{In fact, Pfenning and Davies \cite{logic-modal-reconstruction,logic-cmtt}
never mention comonads explicitly. This is done in later work by Gabbay et al. \cite{logic-cmtt-semantics}, 
but the connection between the language and comonads is not as direct as in case of monadic or
comonadic semantics covered in the last few pages.}.

In languages inspired by modal logics, types can have the form $\square \tau$. In the work of
Pfenning and Davies, this means a term that is provable with no assumptions. In distributed 
programming language ML5, Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus} use the 
$\square \tau$ type to mean \emph{mobile code}, that is code that can be evaluated at any node of a 
distributed system (the evaluation corresponds to the axiom $\square \tau \rightarrow \tau$). 
Finally, Davies and Pfenning \cite{logic-modal-staged} consider staged computations and interpret 
$\square \tau$ as a type of unevaluated expressions of type $\tau$ (with no free variables).

In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variales
of the (unevaluated) expression. To keep the syntax consistent with earlier examples, we use 
$\ctyp{\cclrd{\Psi}}{\tau}$ for a type $\square \tau$ with an annotation $\cclrd{\Psi}$. The type 
is a comonadic counterpart to the \emph{indexed monads} used by Wadler and Thiemann when linking 
monads and effect systems and, indeed, it gives rise to a language that tracks context-dependence 
of computations in a type system.

In staged computation, the type $\ctyp{\cclrd{\Psi}}{\alpha}$ represents an expression 
that requires the context $\Psi$ (i.e.~the expression is an open term that requires variables $\Psi$).
The Figure~\ref{fig:modal-meta} shows two typing rules for such language. The rules directly
correspond to the two operations of a comonad and can be interpreted as follows:

\begin{itemize}
\item (\emph{eval}) corresponds to $\ident{counit} : \ctyp{\cclrd{\emptyset}}{\alpha} \rightarrow \alpha$. 
  It means that we can evaluate a closed (unevaluated) term and obtain a value. Note that the 
  rule requires a specific context annotation (empty set of free variables). 
  It is not possible to evaluate an open term.

\item (\emph{letbox}) corresponds to $\ident{cobind} : (\ctyp{\cclrd{\Psi}}{\alpha} \rightarrow \beta) 
  \rightarrow \ctyp{\cclrd{\Psi}, \clrd{\Phi}}{\alpha} \rightarrow \ctyp{\cclrd{\Phi}}{\beta}$. 
  It means that given a term which requires variable context $\cclrd{\Psi}, \cclrd{\Phi}$ 
  (expression $e_1$) and a function that turns a term needing $\cclrd{\Psi}$ into an evaluated 
  value (expression $e_2$), we can construct a term that requires just $\cclrd{\Phi}$.
\end{itemize}

@@

The annotated \ident{cobind} operation that corresponds to (\emph{letbox}). We discuss this 
in Section~\ref{sec:unified-meta}. An interesting aspect is that it propagates the 
context-requirements ``backwards''. The input expression (second parameter) requires a 
combination of contexts that are required by the two components -- those required by the 
input of the function (first argument) and those required by the resulting expression (result). 
This is another key aspect that distinguishes coeffects from effect systems.

@@

\begin{figure}
\begin{equation*}
\inference[(eval)]
  {\Gamma \vdash e : \ctyp{\cclrd{\emptyset}}{\tau}}
  {\Gamma \vdash !e : \tau}
\end{equation*}
\begin{equation*}
\inference[(letbox)]
  { \Gamma \vdash e_1 : \ctyp{\cclrd{\Phi}, \cclrd{\Psi}}{\tau_1} & 
    \Gamma, x\!:\!\ctyp{\cclrd{\Phi}}{\tau_1} \vdash e_2 : \tau_2 }
  { \Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{\Psi}}{\tau_2}}
\end{equation*}

@@

\paragraph{Thesis perspective}
As mentioned earlier, we are interested in designing context-dependent languages and so we
use comonads as \emph{language semantics}. Uustalu and Vene present a semantics of 
context-dependent computations in terms of comonads. We provide the rest of the story known 
from the marriage of monads and effects. We develop coeffect calculus with a type system that 
tracks the context requirements more precisely (by annotating the types) and we add indexing 
to comonads and link the two by giving a formal semantics.

The \emph{meta-language} approach of Pfenning, Davies and Nanevski et al. is closely related to
our work. Most importantly, Contextual Modal Type Theory (CMTT) uses indexed $\square$ modality
which corresponds to indexed comonads (in a similar way in which effect systems correspond to 
indexed monads). The relation between CMTT and comonads has been suggested by
Gabbay et al. \cite{logic-cmtt-semantics}, but the meta-language employed by CMTT does not 
directly correspond to comonadic operations. For example, our (\emph{letbox}) typing rule from
Figure~\ref{fig:modal-meta} is not a primitive of CMTT and would correspond to 
$\ident{box}(\cclrd{\Psi}, \ident{letbox}(e_1, x, e_2))$. Nevertheless, the indexing in 
CMTT provides a useful hint for adding indexing to the work of Uustalu and Vene.

@@

\section{Through sub-structural and bunched logics}
\label{sec:path-logic}

In the coeffect system for tracking resource usage outlined earlier, we associated additional
contextual information (set of available resources) with the variable context of the typing 
judgement: $\Gamma @ \sigma \vdash e : \alpha$. In other words, our work focuses on ``what is
happening on the left hand side of $\vdash$''.

@@

In this section, we look at type systems that reconsider $\Gamma$ in a number of ways. 
First of all, sub-structural type systems \cite{substruct-attpl-intro} restrict the use of variables
in the language. Most famously linear type systems introduced by Wadler \cite{substruct-linear-change} 
can guarantee that variable is used exactly once. This has interesting implications for memory
management and I/O.

In bunched typing developed by O'Hearn \cite{substruct-bunched}, the variable context is a tree 
formed by multiple different constructors (e.g.~one that allows sharing and one that does not). 
Most importantly, bunched typing has contributed to the development of separation logic
\cite{substruct-separation-logic} (starting a fruitful line of research in software verification), 
but it is also interesting on its own.

@@

\paragraph{Sub-structural type systems}

Traditionally, $\Gamma$ is viewed as a set of assumptions and typing rules admit (or explicitly
include) three operations that manipulate the variable contexts which are shown in 
Figure~\ref{fig:substructural-rules}. The (\emph{exchange}) rule allows us to reorder variables
(which is implicit, when assumptions are treated as set); (\emph{weakening}) makes it possible
to discard an assumption -- this has the implication that a variable may be declared but never
used. Finally, (\emph{contraction}) makes it possible to use a single variable multiple times
(by joining multiple variables into a single one using substitution).

\begin{figure}
\begin{equation*}
\inference[(exchange)]
  {\Gamma, x\!:\!\alpha, y\!:\!\tau_2 \vdash e : \gamma}
  {\Gamma, y\!:\!\tau_2, x\!:\!\alpha \vdash e : \gamma}
\quad
\inference[(weakening)]
  {\Gamma, \Delta \vdash e : \gamma}
  {\Gamma, x\!:\!\alpha, \Delta \vdash e : \gamma}
\end{equation*}
\begin{equation*}
\inference[(contraction)]
  {\Gamma, x\!:\!\alpha, y\!:\!\alpha, \Delta \vdash e : \gamma}
  {\Gamma, x\!:\!\alpha, \Delta \vdash \subst{e}{y}{x} : \gamma}
\end{equation*}

@@

In sub-structural type systems, the assumptions are typically treated as a list. As a result,
they have to be manipulated explicitly. Different systems allow different subset of the rules.
For example, \emph{affine} systems allows exchange and weakening, leading to a system where 
variable may be used at most once; in \emph{linear} systems, only exchange is permitted and so 
every variable has to be used exactly once.

When tracking context-dependent properties associated with individual variables, we need to 
be more explicit in how variables are used. Sub-structural type systems provide a way to do this.
Even when we allow all three operations, we can track which variables are used and how
(and use that to track additional contextual information about variables).

@@

\paragraph{Bunched type systems}
Bunched typing makes one more refinement to how $\Gamma$ is treated. Rather than having a list
of assumptions, the context becomes a tree that contains variable typings (or special identity
values) in the leaves and has multiple different types of nodes. The context can be defined,
for example, as follows:
%
\begin{equation*}
\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep I \sep \Gamma, \Gamma \sep 1 \sep \Gamma; \Gamma
\end{equation*}
%
The values $I$ and $1$ represent two kinds of ``empty'' contexts. More interestingly, non-empty
variable contexts may be constructed using two distinct constructors -- $\Gamma, \Gamma$ and 
$\Gamma; \Gamma$ -- that have different properties. In particular, weakening and contraction is
only allowed for the $;$ constructor, while exchange is allowed for both.

@@

One important note about bunched typing is that it requires a different interpretation. The omission
of weakening and contraction in linear logic means that variable can be used exactly once. 
In bunched typing, variables may still be duplicated, but only using the ``;'' separator.
The type system can be interpreted as specifying whether a variable may be shared between the 
body of a function and the context where a function is declared. The system introduces two 
distinct function types $\alpha \rightarrow \tau_2$ and $\alpha~ \textendash\!\!\!\ast \tau_2$
(corresponding to ``;'' and ``,'' respectively). The key property is that only the first kind
of functions can share variables with the context where a function is declared, while the second
restricts such sharing. We do not attempt to give a detailed description here as it is not 
immediately to coeffects -- for more information, refer to O'Hearn's introduction 
\cite{substruct-bunched}.

\begin{figure}
\begin{equation*}
\inference[(exchange1)]
  {\Gamma(\Delta, \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma, \Delta) \vdash e : \alpha}
\quad
\inference[(weakening)]
  {\Gamma(\Delta) \vdash e : \alpha}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\inference[(exchange2)]
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
\quad
\inference[(contraction)]
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Delta) \vdash \subst{e}{\Sigma}{\Delta} : \alpha}
\end{equation*}
\caption{Exchange, weakening and contraction rules for bunched typing}
\label{fig:substructural-bunched}
\end{figure}

@@

\paragraph{Thesis perspective}

Our work can be viewed as annotating bunches. Such annotations then specify additional information
about the context -- or, more specifically, about the sub-tree of the context. Although this is not
the exact definition used in Chapter~X, we could define contexts as follows:
%
\begin{equation*}
\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep 1 \sep \Gamma, \Gamma \sep \Gamma @ \sigma
\end{equation*}
%
Now we can not only annotate an entire context with some information (as in the simple coeffect
system for tracking resources that used judgements of a form $\Gamma @ \sigma \vdash e : \alpha$).
We can also annotate individual components. For example, a context containing variables $x,y,z$
where only $x$ is used could be written as $(x\!:\!\alpha \,@\, \ident{used}), ((y\!:\!\alpha, z\!:\!\alpha) 
\,@\, \ident{unused})$.

For the purpose of this introduction, we ignore important aspects such as how are nested annotations
interpreted. The main goal is to show that coeffects can be easily viewed as an extension to the 
work on bunched logic. Aside from this principal connection, \emph{structural coeffects} also 
use some of the proof techniques from the work on bunched logics, because they also use tree-like
structure of variable contexts.

@@

The importance of context-aware computations is perhaps most obvious when considering mobile
application, client/server web applications or even the internet of things. A pioneering work
in the area using functional languages has been done by Serrano \cite{app-hop-diffuse,app-hop-lang}
(which also inspired the example presented in Chapter~\ref{ch:intro}). His HOP language supports 
cross-compilation and programs execute in different contexts. However, HOP is not statically 
type checked.

@@

We approach the problem from a different perspective, building on the tradition of 
statically-typed functional programming languages and their theories.

@@

This chapter presented four different pathways leading to the idea of coeffects. We also 
introduced the most important related work, although presenting related work was not the
main goal of the chapter. The main goal was to show the idea of coeffects as a logical follow up
to a number of research directions. For this reason, we highlighted only certain aspects of 
related work -- the remaining aspects as well as important technical details are covered 
in later chapters.

The first pathway looks at applications and systems that involve notion of \emph{context}.
The two coeffect calculi we present aim to unify some of these systems. The second pathway
follows as a dualization of well-known effect systems. However, this is not simply a syntactic
transformation, because coeffect systems treat lambda abstraction differently. The third 
pathway follows by extending comonadic semantics of context-dependent computations with 
indexing and building a type system analogous to effect system from the ``marriage of effects
and monads''. Finally, the fourth pathway starts with sub-structural type systems. Coeffect
systems naturally arise by annotating bunches in bunched logics with additional information.

% \section{Missing}
% ~
% 
% indexed/layered/etc. monads, productors or whatever

>>>>

\noindent
Without providing much details, we note that well known examples of monads include the partiality
monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in 
Haskell, list monad ($\mtyp{}{\tau} = \mu \gamma.1 + (\tau \times \gamma)$) and other.
In programming language semantics, monads can be used in two distinct ways.

@@

In this thesis, we generally follow the first approach -- this means that we work with an existing
programming language without needing to add additional constructs corresponding to the primitives
of our semantics (the alternative is discussed in Section~\ref{sec:unified-meta}). To clarify the 
difference, the following two sections show a minimal example of both formal systems. We follow 
Moggi and start with language where judgements have the form $x\!:\!\tau_1 \vdash e : \tau_2$ with 
exactly one variable\footnote{This simplifies the examples as we do not need \emph{strong} monad, 
but that is an orthogonal issue to the distinction between language semantics and meta-language.}.

@@

\paragraph{Language semantics.}
To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
$\ctyp{}{\tau} = \mu \gamma.\tau + (\tau \times \gamma)$. A value of the type is either
the last element $\tau$ or an element followed by another non-empty list $\tau \times \gamma$
(consisting of the head $\tau$ and the tail $\gamma$). Note that the list must be non-empty, 
otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In 
the following, we write $(l_1, \ldots, l_n)$ for a list of $n$ elements:
%
\begin{equation*}
\begin{array}{rcl}
\ident{counit}~(l_1, \ldots, l_n) &=& l_1\\
\ident{cobind}~f~(l_1, \ldots, l_n) &=& (f (l_1, \ldots, l_n), f (l_2, \ldots, l_n), \ldots, f (l_n))
\end{array}
\end{equation*}
%
The \ident{counit} operation returns the current (first) element of the (non-empty) list.
The \ident{cobind} operation creates a new list by applying the context-dependent function $f$
to the entire list, to the suffix of the list, to the suffix of the suffix and so on. Interestingly,
it preserves the \emph{shape} of the list as it turns a list of $n$ elements into another list
of $n$ elements.

@@

\paragraph{Meta-language interpretation.} ~To briefly demonstrate the app\-roach that employs comonads
as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
Nanevski. We do not attempt to provide a precise overview of their work. The main purpose 
of the following discussion is to provide a different intuition behind comonads, and to present an 
example of a language that includes comonad as a type constructor, together with language primitives
corresponding to comonadic operations\footnote{In fact, Pfenning, Davies and Nanevski 
\cite{logic-modal-reconstruction,logic-cmtt} never mention comonads explicitly. This is done in later 
work by Gabbay et al. \cite{logic-cmtt-semantics},  but the connection between the language and comonads
is not as direct as in case of monadic or comonadic semantics covered in the previous section.}.

In languages inspired by modal logics, types can have the form $\square \tau$. In the work of
Pfenning and Davies, this is the type of a term that is provable with no assumptions. In distributed 
programming language ML5 by Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus}, the 
$\square \tau$ type means \emph{mobile code}, that is code that can be evaluated at any node of a 
distributed system (the evaluation corresponds to the axiom $\square \tau \rightarrow \tau$). 
Finally, Davies and Pfenning \cite{logic-modal-staged} consider staged computations and interpret 
$\square \tau$ as a type of unevaluated expressions of type $\tau$ (with no free variables).

In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variales
of the (unevaluated) expression. We write $\square^{\cclrd{\Psi}} \tau$ for a type of expressions
that requires a context $\Psi$. The type is a comonadic counterpart to \emph{indexed monads} used by 
Wadler and Thiemann when linking monads and effect systems and, indeed, it gives rise to a language 
that tracks context-dependence of computations in a type system.

In staged computation, the type $\ctyp{\cclrd{\Psi}}{\tau}$ represents an expression 
that requires the context $\Psi$ (i.e.~the expression is an open term that requires variables $\Psi$).
The Figure~\ref{fig:modal-meta} shows two typing rules for such language. The rules directly
correspond to the two operations of a comonad and can be interpreted as follows:

\begin{itemize}
\item (\emph{eval}) corresponds to $\ident{counit} : \ctyp{\cclrd{\emptyset}}{\alpha} \rightarrow \alpha$. 
  It indicates that we can evaluate a closed (unevaluated) term and obtain a value. Interestingly, the 
  rule requires a specific context annotation (empty set of free variables). 
  It is not possible to evaluate an open term.

\item (\emph{letbox}) corresponds to $\ident{cobind} : (\ctyp{\cclrd{\Psi}}{\alpha} \rightarrow \beta) 
  \rightarrow \ctyp{\cclrd{\Psi}, \cclrd{\Phi}}{\alpha} \rightarrow \ctyp{\cclrd{\Phi}}{\beta}$. 
  Given a term which requires variable context $\cclrd{\Psi}, \cclrd{\Phi}$ 
  (expression $e_1$) and a function that turns a term needing $\cclrd{\Psi}$ into an evaluated 
  value (expression $e_2$), we can construct a term that requires just $\cclrd{\Phi}$.
\end{itemize}

@@

The annotated \ident{cobind} operation that corresponds to (\emph{letbox}). An interesting aspect 
is that it propagates the context-requirements ``backwards''. The input expression (second parameter) 
requires a combination of contexts that are required by the two components -- those required by the 
input of the function (first argument) and those required by the resulting expression (result). 
This is another key aspect that distinguishes coeffects from effect systems. We return back to 
the meta-language approach of embedding comonads in Section~\ref{sec:unified-meta}.

@@

\begin{figure}
\begin{equation*}
\inference[(eval)]
  {\Gamma \vdash e : \square^{\cclrd{\emptyset}}{\tau}}
  {\Gamma \vdash !e : \tau}
\end{equation*}
\begin{equation*}
\inference[(letbox)]
  { \Gamma \vdash e_1 : \square^{\cclrd{\Phi}, \cclrd{\Psi}}{\tau_1} & 
    \Gamma, x\!:\!\square^{\cclrd{\Phi}}{\tau_1} \vdash e_2 : \tau_2 }
  { \Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \square^{\cclrd{\Psi}}{\tau_2}}
\end{equation*}

@@

\paragraph{Thesis perspective.}
As mentioned earlier, we are interested in designing context-dependent languages and so we
use comonads for \emph{language semantics}. Uustalu and Vene present a semantics of 
context-dependent computations in terms of comonads. We provide the rest of the story known 
from the marriage of monads and effects. We develop coeffect calculus with an type system that 
tracks the context requirements more precisely (by annotating the types) and we add indexing 
to comonads and link the two by giving a formal semantics. The indexing allows us to capture 
applications that do not fit into the model provided by plain comonads.

The \emph{meta-language} approach of Pfenning, Davies and Nanevski is closely related to
our work. Most importantly, Contextual Modal Type Theory (CMTT) uses indexed $\square$ modality
which corresponds to indexed comonads (in a similar way in which effect systems correspond to 
indexed monads). The relation between CMTT and comonads has been suggested by
Gabbay et al. \cite{logic-cmtt-semantics}, but the meta-language employed by CMTT does not 
directly correspond to comonadic operations. For example, our (\emph{letbox}) typing rule from
Figure~\ref{fig:modal-meta} is not a primitive of CMTT and would correspond to 
$\ident{box}(\cclrd{\Psi}, \ident{letbox}(e_1, x, e_2))$. Nevertheless, the indexing in 
CMTT provides a useful hint for adding indexing to the work of Uustalu and Vene.

@@

\section{Through substructural and bunched logics}
\label{sec:path-logic}

In the coeffect system for tracking resource usage outlined earlier, we associated additional
contextual information (set of available resources) with the variable context of the typing 
judgement: $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$. In other words, our work focuses on 
``what is happening on the left hand side of $\vdash$''.

@@

In this section, we look at type systems that reconsider $\Gamma$ in a number of ways. 
First of all, substructural type systems \cite{substruct-attpl-intro} restrict the use of variables
in the language. Most famously linear type systems introduced by Wadler \cite{substruct-linear-change} 
can guarantee that a variable is used exactly once. This has interesting implications for memory
management and I/O.

In bunched typing developed by O'Hearn \cite{substruct-bunched}, the variable context is a tree 
formed by multiple different constructors (e.g.~one that allows sharing and one that does not). 
Most famously, bunched typing has contributed to the development of separation logic
\cite{substruct-separation-logic} (starting a fruitful line of research in software verification), 
but it is also interesting on its own.

@@

\subsection{Substructural type systems.}

Traditionally, $\Gamma$ is viewed as a set of assumptions and typing rules admit (or explicitly
include) three transformations that manipulate the variable contexts which are shown in 
Figure~\ref{fig:substructural-rules}. The (\emph{exchange}) rule allows reordering of variables
(which is implicit when assumptions are treated as set); (\emph{weakening}) makes it possible
to discard an assumption -- this has the implication that a variable may be declared but never
used. Finally, (\emph{contraction}) makes it possible to use a single variable multiple times
(in the rule, this is done explicitly by joining multiple variables into a single one using 
substitution).

In substructural type systems, the assumptions are typically treated as a list. As a result,
they have to be manipulated explicitly. Different systems allow different subsets of the rules.
For example, \emph{affine} systems allows exchange and weakening, leading to a system where 
variable may be used at most once; in \emph{linear} systems, only exchange is permitted and so 
every variable has to be used exactly once.

When tracking context-dependent properties associated with individual variables, we need to 
be more explicit in how variables are used. Substructural type systems provide a way to do this.
Even if we allow all three operations, we can use a variation on the three rules (exchange, 
weakening and contraction) to track which variables are used and how (and to track additional 
contextual information about variables).

\begin{figure}
\begin{equation*}
\tyrule{exchange}
  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_2 \vdash e : \gamma}
  {\Gamma, y\!:\!\tau_2, x\!:\!\tau_1 \vdash e : \gamma}
\end{equation*}
\begin{equation*}
\tyrule{weakening}
  {\Gamma, \Delta \vdash e : \gamma}
  {\Gamma, x\!:\!\tau, \Delta \vdash e : \gamma}
\end{equation*}
\begin{equation*}
\tyrule{contraction}
  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_1, \Delta \vdash e : \tau_2}
  {\Gamma, x\!:\!\tau_1, \Delta \vdash \subst{e}{y}{x} : \tau_2}
\end{equation*}

@@

@@

\subsection{Bunched type systems.}
Bunched typing makes one more refinement to how $\Gamma$ is treated. Rather than having a list
of assumptions, the context becomes a tree that contains variable typings (or special identity
values) in the leaves and has multiple different types of nodes. The context can be defined,
for example, as follows:
%
\begin{equation*}
\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep I \sep \Gamma, \Gamma \sep 1 \sep \Gamma; \Gamma
\end{equation*}
%
The values $I$ and $1$ represent two kinds of ``empty'' contexts. More interestingly, non-empty
variable contexts may be constructed using two distinct constructors -- $\Gamma, \Gamma$ and 
$\Gamma; \Gamma$ -- that have different properties. In particular, weakening and contraction is
only allowed for the $;$ constructor, while exchange is allowed for both.

@@

One important note about bunched typing is that it requires a different interpretation. The omission
of weakening and contraction in linear logic means that variable can be used exactly once. 
In bunched typing, variables may still be duplicated, but only using the ``;'' separator.
The type system can be interpreted as specifying whether a variable may be shared between the 
body of a function and the context where a function is declared.

The system introduces two 
distinct function types $\tau_1 \rightarrow \tau_2$ and $\tau_1~ \textendash\!\!\!\ast \tau_2$
(corresponding to ``;'' and ``,'' respectively). The key property is that only the first kind
of functions can share variables with the context where a function is declared, while the second
restricts such sharing. We do not attempt to give a detailed description here as it is not 
immediately related to coeffects -- for more information, refer to O'Hearn's introduction 
\cite{substruct-bunched}.

\begin{figure}
\begin{equation*}
\tyrule{exchange1}
  {\Gamma(\Delta, \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma, \Delta) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{exchange2}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{weakening}
  {\Gamma(\Delta) \vdash e : \alpha}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{contraction}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Delta) \vdash \subst{e}{\Sigma}{\Delta} : \alpha}
\end{equation*}
\caption{Exchange, weakening and contraction rules for bunched typing}
\label{fig:substructural-bunched}
\end{figure}

@@

\paragraph{Thesis perspective.}

From the perspective of substructural and bunched types, our work can be viewed as annotating 
bunches. Such annotations then specify additional information about the context -- or, more
specifically, about the sub-tree of the context. Although this is not the exact definition used in 
Chapter~\ref{ch:unified}, we could define contexts as follows:
%
\begin{equation*}
\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep 1 \sep \Gamma, \Gamma \sep \coctx{\Gamma}{\cclrd{r}}
\end{equation*}
%
Now we can not only annotate an entire context with some information (as in the simple coeffect
system for tracking resources that used judgements of a form $\cclrd{\Gamma}{\cclrd{r}} \vdash e : \tau$).
We can also annotate individual components. For example, a context containing variables $x,y,z$
where only $x$ is used could be written as $\coctx{(x\!:\!\tau_1}{\ident{\cclrd{used}}}, \coctx{(y\!:\!\tau_2, z\!:\!\tau_3)}{\ident{\cclrd{unused}}}$.

For the purpose of this introduction, we ignore important aspects such as how are nested annotations
interpreted. The main goal is to show that coeffects can be easily viewed as an extension to the 
work on bunched logic. Aside from this principal connection, \emph{structural coeffects} also 
use some of the proof techniques from the work on bunched logics.

@@

The importance of context-aware computations is perhaps most obvious when considering mobile
application, client/server web applications or even the internet of things. A pioneering work
in the area using functional languages has been done by Serrano \cite{app-hop-diffuse,app-hop-lang}
(which also inspired the motivating example presented in Chapter~\ref{ch:intro}). His HOP language supports 
cross-compilation and programs execute in different contexts. However, HOP is not statically 
type checked.

@@

We approach the problem from a different perspective, building on the tradition of 
statically-typed functional programming languages, focusing on type systems as the primary way
of capturing contextual properties.

@@

This chapter presented three different pathways leading to the idea of coeffects. We also 
introduced the most important related work, although presenting related work was not the
primary goal of the chapter. The primary goal was to present the idea of coeffects as a logical 
follow up to a number of research directions. For this reason, we highlighted only certain aspects 
of the discussed related work -- the remaining aspects as well as important technical details are
covered throughout the thesis.

The first pathway follows as a dualization of well-known effect systems. However, this is not simply 
a syntactic transformation. As we further discuss in the next chapter, coeffect systems treat lambda 
abstraction differently. The second pathway follows by extending comonadic semantics of context-dependent 
computations with indexing and building a type system analogous to effect system from the ``marriage of 
effects and monads''. Finally, the third pathway starts with substructural type systems. Coeffect
systems naturally arise by annotating bunches in bunched logics with additional information. In this
thesis, we mostly follow the first two approaches.
